### 2004-07-15 Documentation/part2.tex
*** empty log message ***

The description of the neutrino oscillation physics is, at least numerically, relatively simple. We use the {\em evolution operator method} to compute the neutrino oscillation probabilities and divide the matter density profile into layers with constant matter densities. For each of these layers, the Hamiltonian in matter is diagonalized in order to propagate the neutrino transition amplitudes. Finally, the transition probability is obtained by the absolute square of the total neutrino transition amplitudes. Depending on the precision of the studied experiment, this approach turns out to be precise enough in Earth matter even for a small number matter density steps. Since we allow an uncertainty of the matter density profile, it is, in fact, in most cases sufficient to consider only one density step with the average matter density together with a matter density uncertainty~\cite{Ohlsson:2003ip}. Note that this approach may not be applicable to quickly varying extraterrestrial matter density profiles.

@@

This picture can also be formulated in a more mathematical way. Let us define $x$ as the true parameter value and $x'$ as the reconstructed parameter value. Similarly, $f(x)$ is the distribution of true parameters values and $p(x')$ is the distribution of reconstructed parameter values. Then the detector function  $D(x,x')$, which describes the mapping performed by the detector, is given by
\begin{eqnarray}
\label{equ:mapping}
p(x')&=&\int dx\, f(x)\cdot D(x,x')\,.
\end{eqnarray}
Obviously \equ{mapping} only describes the detector properly
if the linearity condition is fulfilled. Within this model, a detector
is completely specified by a set of $D(E,E')$ for the energy variable $E$,
and a set $D(F,F')$ for the flavor variable $F$. In general, $D(E,E')$ also depends on the true flavor $F$, as well as $D(F,F')$ depends on the true energy $E$. These sets of mapping functions usually are obtained from a 
full detector simulation and can be obtained by using as input 
distribution $f(x)$ a delta distribution $\delta(x-x_0)$.

@@

The channels are the building blocks for the so-called ``rules''.
\index{norm}{Rule} In general, a rule consists of one or more ``signal'' and ``background'' oscillation channels, which are normalized with efficiencies
(\cf, \figu{rule}). The event numbers from these channels are added {\em before} the $\Delta \chi^2$-value is calculated.\footnote{Note that
in this manual, we $\chi^2$ and $\Delta \chi^2$ are equal, since
for simulated data $\Delta \chi^2 = 0$ at the best-fit point. Thus, we
are using $\chi^2$ and $\Delta \chi^2$ as equal quantities.} In addition, each rule implements an {\em independent} systematics, such as signal and background normalization errors. Eventually, each rule gives a $\Delta \chi^2$-value, and the total $\Delta \chi^2$ of one experiment is obtained by adding the $\Delta \chi^2$'s of all rules (\cf, \figu{experiment}). 
 An example for a rule could look like this: We want to detect electron
  neutrino appearance (``signal''), where the overall efficiency for 
  quasi-elastics electron neutrino events is $0.4$. There is a fraction of 
  $0.01$ of all neutral current events which are mis-identified as 
  quasi-elastic electron neutrino events (``background''). The neutral 
  current fraction is only known within $10\%$ (``background uncertainty'') 
  and there is
an energy scale uncertainty of $100\,\mathrm{MeV}$ (``energy calibration error'').
All this systematics is independent of the other rules.  Thus, a rule connects the event rates to the calculation of a $\Delta \chi^2$ which properly includes systematical errors. The resulting $\Delta \chi^2$ is then the starting point for the oscillation physics analysis. Note again that
\begin{itemize}
\item
 Within each rule the event numbers are added.
\item
 Within each rule the systematics is treated independently from the other rules.
\item
 For each rule the $\Delta \chi^2$ is computed; the $\Delta \chi^2$'s from all rules are added.
\end{itemize}

@@

Experiments are in \GLOBES\ defined by the Abstract Experiment Definition Language (AEDL). The experiment definition is written into a text file using the AEDL syntax. Currently, a number of pre-defined experiment definition files are provided with \GLOBES , which have to be modified manually in order to define new experiments. An IDE (Integrated Developement Environment) for AEDL files is currently in preparation. The application software then uses this text file to initialize the experiment, where other secondary files might read for source fluxes, cross sections \etc . In this section, we show the definition of a very simple neutrino factory in AEDL, where we do not go into details. In the next chapter, we will discuss each of the individual steps in detail.

@@

Of course, the baseline and matter profile have to be specified, too:

@@

{\tt \$baseline = 3000.0}\\

@@

{\tt \$density\_error = 0.05}

@@

one element, because we only use one density layer: We initialize a baseline length of $3000 \, \mathrm{km}$ with a constant matter density of $3.5 \, \mathrm{g/cm^3}$ and a matter density uncertainty of $5\%$ ($1 \sigma$ relative error). Note that the elements in {\tt \$lengthtab} have to add up to the baseline length. 

@@

The first element is the name of the flux, which we have already defined above as defined above. The second element $\pm$ determines whether 

@@

We now give a short introduction to the syntax of AEDL.
 The first eight characters have to be {\tt \%!GLoBES}
in order to avoid parsing megabytes of chunk
 and producing thousands of error messages. 
%
Comments can be used such as in C:
\begin{quote}
{\tt /* This starts a comment\\
 and here the comment ends */
}
\end{quote}
There are pre-defined variables which all start with {\tt \$}. Their range
is also checked. For example,  {\tt 
\$bins} can be only between $0$ and $500$~\footnote{The upper limit is 
there only for safety reasons, the memory is allocated dynamically.}. If one uses a {\tt float} quantity where  an {\tt int} is expected, the {\tt float} will be converted to an {\tt int} in the same way as in C.  For example, we have scalar variables
\begin{quote}
{\tt
\$bins = 10\\
\$baseline = 1200.0
}
\end{quote}
and simple lists
\begin{quote}
{\tt
\$densitytab=\{1.0,2.2343,3.3432\} 
}
\end{quote}
%
Since there are often groups of data which we want to refer to later,
environments can be used. This is illustrated 
with the channel definition part:
\begin{quote}
{\tt channel(\#ch1)<\\
\tb  $\ldots$\\
>
}
\end{quote}
The first part is the type of environment, which is {\tt channel} here. 
There are the following types of environment in AEDL:
\begin{quote}
{\tt flux\\
cross\\
channel\\
energy\\
rule
}
\end{quote}
Besides the environment type, there is a user-defined name 
beginning with {\tt \#}
in the above example: {\tt \#ch1}. It can be used later to refer to the 
channel defined in {\tt <$\ldots$>}. Those names are so-called 
``automatic variables'' and have to start with {\tt \#}. Note that these names have to be unique and can only be refered to after their definition.
However, similar to C, one can give a header definition before:
\begin{quote}
{\tt    channel(\#ch2)<>}
\end{quote}
Now one can refer to the name {\tt \#ch2}, while the actual channel definition comes later. The internal representation of this automatic
variable is a number, which obtains its value from a counter for each type of environment. For example, for {\tt channel} the counter is {\tt numofchannels}. The counter keeps track of how many different names 
there are for one type of environment, which means that it counts the number of channels, rules, energy resolution functions \etc . Thus, the automatic
variables are numbered in the order of their definition, and the number
can later be used to refer to them in the C code (from $0$ to {\tt numof...}$-1$). Within each environment type, there are several 
variables beginning with {\tt @}, which can only be used within the 
appropriate type of environment. In many cases, 
they have a special syntax, such as {\tt @channel}.

\index{aedl}{NEXT@{\tt \#NEXT\#}}
If you want to have several experiments in one file, separate the different
 experiments by 
\begin{quote}
{\tt    \#NEXT\#}
\end{quote}
This command resets the counters for channels, rules and energy resolution
functions. All variables have their scope limited 
by either {\tt \%!GLoBES, \#NEXT\#} or {\tt EOF}.  This allows 
a consistent treatment of various experiments in one file.

\index{aedl}{inlcude@{\tt include}}
As another feature of AEDL one can use include files with the {\tt include} command. Includes can be nested up to {\tt MAX\_INCLUSION\_DEPTH}, which is currently set to $10$. Error reporting works 
 for nested includes, too. The included file is not required to begin 
 with {\tt \%!GLoBES} to facilitate cut and paste:
\begin{quote}
{\tt include "./file\_1"}
\end{quote}
With this include mechanism, one can use constructions such as 
\begin{quote}
{\tt    include  "NuFact.gls"\\
        \#NEXT\#\\
        include   "JHFHK.gls"
}
\end{quote}
in order to initialize a combined analysis of the experiments defined in the files {\tt NuFact.gls} and {\tt JHFHK.gls}. As one sees from the example above
file names are always quoted in \AEDL.\index{norm}{File names}
Even if one uses the
automatic variable {\tt \#CC} in both experiments, 
but the cross section data are different (for example, because of different target nuclei), the correct 
cross section data will be applied to each of the experiments. 
Note that, alternatively, one can 
also load both files successively.

Furthermore, one can define constants such as
\begin{quote}
{\tt
Pi = 3.141
}
\end{quote}
These constants can not only be defined within one \AEDL\ file but also
by the calling C program, which allows to use a simple but powerful variable
substitution mechanism as described in~\Sec~\ref{sec:aedlparams}.
\index{norm}{External \AEDL\ parameters}

Alos some simple algebraic manipulations are possible, such as
\begin{quote}
{\tt
Pi+1\\
\verb+Pi^2+\\
sin(Pi/2)\\
}
\end{quote}
The following mathematical functions from {\tt <math.h>} are available: 
{\tt sin}, {\tt cos}, {\tt tan}, {\tt asin}, {\tt acos}, {\tt atan}, 
{\tt log}, {\tt log10}, {\tt exp}, {\tt sqrt}.
\index{aedl}{sin@{\tt sin}}
\index{aedl}{cos@{\tt cos}}
\index{aedl}{tan@{\tt tan}}
\index{aedl}{asin@{\tt asin}}
\index{aedl}{acos@{\tt acos}}
\index{aedl}{atan@{\tt atan}}
\index{aedl}{log@{\tt log}}
\index{aedl}{log10@{\tt log10}}
\index{aedl}{exp@{\tt exp}}
\index{aedl}{sqrt@{\tt sqrt}}

@@

\begin{table}[t!]
\begin{tabular}{|clp{7cm}|}
\hline
{\tt \$profiletype} & Additional variables & Description \\ 
\hline
$1$ & - & Average density (constant) \\
$2$ & {\tt \$densitysteps} & PREM profile with given number of equidistant steps \\
$3$ & {\tt \$lengthtab},  {\tt \$densitytab} & Arbitrary profile (table of layer thicknesses, table of densities) \\
\hline
\end{tabular}
\caption{\label{tab:profiletypes} Different matter density profiles which can be used with \GLOBES .}
\end{table}

@@

Furthermore, the matter density profile along the baseline
has to be specified. The simplest matter density profile is a constant matter density profile with the average matter density from the PREM\cite{Stacey} onion shell model of the Earth\index{norm}{PREM}\index{norm}{Earth matter density}:
\begin{quote}
{\tt \$profiletype=1 }
\end{quote}
%
For a better approximation of the realistic Earth matter density profile, one can use an arbitrary number of equidistant steps of the PREM profile:
\begin{quote}
{\tt \$profiletype=2 } \\
{\tt \$densitysteps=20 }
\end{quote}
Note that the value of {\tt \$densitysteps} is time-critical, since the
computation time of oscillation probabilities is directly 
proportional to the number of layers.
%
As a  third possibility, one can specify the matter density profile 
manually with a list of thicknesses and densities of the matter density layers. This example uses two density steps with two different densities:
\begin{quote}
{\tt \$profiletype=3 } \\
{\tt \$densitytab=\{2.8, 3.5\}}\\
{\tt \$lengthtab=\{1000.0, 2000.0\}}\\
\end{quote}
It is important that both lists habe the same length and that the  thicknesses given in  {\tt \$lengthtab} add up to the value of
{\tt \$baseline}. In addition, matter densities are always given in $g/cm^3$.
%
This approach can also be used for a constant matter density profile with
a specific matter density:
\begin{quote}
{\tt \$profiletype=3 } \\
{\tt \$densitytab=\{3.5\}}\\
{\tt \$lengthtab=\{3000.0\}}\\
\end{quote}
The possible options for matter density profiles are summarized in \Tab~\ref{tab:profiletypes}.

@@

A simple channel definition in \GLOBES\ consists of the flux,
the CP-sign of the initial state, the initial flavor, the final flavor,
the cross sections, and the energy resolution function. In order to refer to
flux, cross sections, and energy resolution functions, they have to be 
defined before with their {\tt \#name} in the respective environments. 
Thus, a simple definition of a channel\index{aedl}{{\tt channel}} is
\begin{quote}
{\tt channel(\#channel\_1)<\\
\tb @channel = \#flux : $+$: muon: muon: \#cross: \#energy\\
>}
\end{quote}

\index{norm}{Oscillation!switching off}
It is also possible to define a channel as no-oscillation by using the
prefix {\tt NOSC\_}\index{aedl}{{\tt channel}!{\tt NOSC\_}} 
in either the initial  flavour or 
the final flavour, like this
\begin{quote}
{\tt channel(\#channel\_1)<\\
\tb @channel = \#flux : $+$: NOSC\_muon: muon: \#cross: \#energy\\
>}
\end{quote}

 In that case
all diagonal probabilities $P_{\alpha\alpha}$ are unity and all off-diagonal
probablities  $P_{\alpha\beta}$ are zero. This is \eg\ especially useful for
neutral current events since they do not depend on any oscillation 
parameters\footnote{at least in the absence of sterile neutrinos}. The channels
marked as {\tt NOSC\_} are only computed by a call to \GLB{SetRates}
and not recomputed in the subsequent fit (which calls \GLB{SetNewRates}).
Therefore this feature can be used to speed up the rate 
computation considerably, especially in cases where a large set
of channels exist which are only used for the computation of backgrounds.
Usually it is an excellent approximation to treat backgrounds as if they
would not be affected by oscillations.

@@

In this section, we first discuss the principles of the energy smearing, where it is assumed that the reader is familiar with \Sec~\ref{sec:channel}. Then we introduce an automatic energy smearing algorithm, which is fairly simple to understand and applicable to most beam-based experiments. 
In most cases, the reader may want to continue to the next section after reading these two subsections. In the third subsection, we describe a more elaborate (and slower) smearing algorithm, which can be used together with rather fast neutrino oscillations compared to the bin size, such as for reactor experiments. Eventually, we show how one can use a manual smearing matrix instead of using one of the implemented algorithms.

@@

Before we come to the calculation algorithms, it is useful to understand
the general evaluation algorithm. As it is illustrated in \figu{energies}, 
\GLOBES\ uses several levels with respect to the energy ranges:
\begin{description}
\item[Sampling point level]
 This level is used internally to evaluate the integrand in \eq~(\ref{eq:simple_int}) at all sampling points. The energy scale is the actual incident neutrino energy $E$. Note that the sampling range should exceed the energy bin range by about the energy resolution error, since the corresponding sampling points have to be evaluated by the algorithm. 
However, this exceeding is only
necessary to avoid aliasing effects, if the integrand in \eq~(\ref{eq:simple_int}) is large at the limits of the sampling range.
For a manual definition of the sampling points (depends on algorithm), use
\index{aedl}{sampling points@{\tt \$sampling\_points}}
\index{aedl}{sampling min@{\tt \$sampling\_min}}
\index{aedl}{sampling max@{\tt \$sampling\_max}}
\begin{quote}
{\tt
\$sampling\_points = 20\\
\$sampling\_min =          4.0\\
\$sampling\_max =         50.0
}
\end{quote}
for equidistant sampling points. Arbitrarily spaced sampling points can 
be specified with 
{\tt \$sampling\_stepsize}\index{aedl}{sampling stepsize@{\tt \$sampling\_stepsize}}
\begin{quote}
{\tt
\$sampling\_stepsize=\{1.0,2.0,3.0,4.0,5.0,...\}
}
\end{quote}

@@

In order to ensure that fast oscillating probabilities do not lead to 
aliasing\index{norm}{Aliasing}
it is desirable to impose a low-pass filter already during the calculation
of the probabilities itself. This is possible and implemented has a highly
experimental feature called ``filter''\index{norm}{Filter}. 
The calculation of oscillation
probabilities basically is a computation of phase differences. Restricting
the maximum admissible size of those phase difference effectively filters
the high frequency component of the oscillation probability. This idea is
implemented according to
\begin{eqnarray}
\label{eq:filter_a}
P_{\alpha\beta}(E)&=&\sum_{ij}
U_{\alpha j} U^*_{\beta j} U^*_{\alpha i} U_{\beta i} 
e^{-i\Phi_{ij}}\times 
e^{ -\Phi_{ij}^2/\sigma_f(E)^2 }\,,
\end{eqnarray}
where $\Phi_{ij}:=\Delta m_{ij}^2 L/2E$ is the usual phase difference and
the last term is a Gau\ss ian filter with width $\sigma_f(E)$. Choosing
$\sigma_f(E):=\sigma_f^0 \cdot E$ ensures that this filter behaves 
approximately like an energy resolution function with constant width 
$\sigma_e=\sqrt{2}/\sigma_f^0$, \ie\
\begin{equation}
\label{eq:filter_b}
\int d\tilde E\quad P(\tilde E) \frac{1}{\sigma_e\,\sqrt{2\pi}}\,
e^{-\frac{(E-\tilde E)^2}{2\sigma^2_e}}\,.
\end{equation}
The equivalence of equation~\ref{eq:filter_a} and equation~\ref{eq:filter_b}
is not obvious and connected to the properties of $P_{\alpha\beta}$, 
see~\cite{Kiers:1996zj,Giunti:2003ax}. This feature works \emph{only} 
for vacuum and constant densities and is controlled
by the filer state variable and $\sigma_e$ is set by the filter value 
variable:
\index{aedl}{filter state@{\tt \$filter\_state}}
\index{aedl}{filter value@{\tt \$filter\_value}}
\begin{quote}
{\tt
\$filter\_state = 1\\
\$filter\_value = 2.0\\
}
\end{quote}
would switch the filter feature on and set the width to $2.0\,\mathrm{GeV}$.
The setting of {\tt \$filter\_state} is ignored whenever a density profile
with more than one layer is used.

>>>>

The description of the neutrino oscillation physics is, at least numerically, relatively simple. We use the {\em evolution operator method} to compute the neutrino oscillation probabilities and divide the matter density profile into layers with constant matter densities. For each of these layers, the Hamiltonian in matter is diagonalized in order to propagate the neutrino transition amplitudes. Finally, the transition probability is obtained by the absolute square of the total neutrino transition amplitudes. Depending on the precision of the studied experiment, this approach turns out to be precise enough in Earth matter even if only a small number matter density steps is used. Since we allow an uncertainty of the matter density profile, it is, in fact, in most cases sufficient to consider only one density step with the average matter density together with a matter density uncertainty~\cite{Ohlsson:2003ip}. Note that this approach may not be applicable to quickly varying extraterrestrial matter density profiles.

@@

This picture can also be formulated in a more mathematical way. Let us define $x$ as the true parameter value and $x'$ as the reconstructed parameter value. Similarly, $f(x)$ is the distribution of true parameters values and $p(x')$ is the distribution of reconstructed parameter values. Then the detector function  $D(x,x')$, which describes the mapping performed by the detector, is given by
\begin{eqnarray}
\label{equ:mapping}
p(x')&=&\int dx\, f(x)\cdot D(x,x')\,.
\end{eqnarray}
Obviously \equ{mapping} only describes the detector properly
if the linearity condition is fulfilled. Within this model, a detector
is completely specified by a set of $D(E,E')$ for the energy variable $E$,
and a set $D(F,F')$ for the flavor variable $F$. In general, $D(E,E',F)$ also depends on the incident neutrino flavor $F$, as well as $D(F,F',E)$ depends on the incident neutrino energy $E$. These sets of mapping functions usually are obtained from a 
full detector simulation and can be obtained by using as input 
distribution $f(x)$ a delta distribution $\delta(x-x_0)$.

@@

The channels are the building blocks for the so-called ``rules''.
\index{norm}{Rule} In general, a rule consists of one or more ``signal'' and ``background'' oscillation channels, which are normalized with efficiencies
(\cf, \figu{rule}). The event numbers from these channels are added {\em before} the $\Delta \chi^2$-value is calculated.\footnote{Note that
in this manual, the $\chi^2$ and $\Delta \chi^2$ are equal, since
for simulated data $\Delta \chi^2 = 0$ at the best-fit point. Thus, we
are using $\chi^2$ and $\Delta \chi^2$ as equal quantities.} In addition, each rule implements an {\em independent} systematics, such as signal and background normalization errors. Eventually, each rule gives a $\Delta \chi^2$-value, and the total $\Delta \chi^2$ of one experiment is obtained by adding the $\Delta \chi^2$'s of all rules (\cf, \figu{experiment}). 
 An example for a rule could look like this: We want to detect electron
  neutrino appearance (``signal''), where the overall efficiency for 
  quasi-elastics electron neutrino events is $0.4$. There is a fraction of 
  $0.01$ of all neutral current events which are mis-identified as 
  quasi-elastic electron neutrino events (``background''). The neutral 
  current fraction is only known within $10\%$ (``background uncertainty'') 
  and there is
an energy scale uncertainty of $100\,\mathrm{MeV}$ (``energy calibration error'').
All this systematics is independent of the other rules.  Thus, a rule connects the event rates to the calculation of a $\Delta \chi^2$ which properly includes systematical errors. The resulting $\Delta \chi^2$ is then the starting point for the oscillation physics analysis. Note again that
\begin{itemize}
\item
 Within each rule the event numbers are added.
\item
 Within each rule the systematics is treated independently from the other rules.
\item
 For each rule the $\Delta \chi^2$ is computed; the $\Delta \chi^2$'s from all rules are added.
\end{itemize}

@@

Experiments are in \GLOBES\ defined by the Abstract Experiment Definition Language (AEDL). The experiment definition is written into a text file using the AEDL syntax. Currently, a number of pre-defined experiment definition files are provided with \GLOBES , which have to be modified manually in order to define new experiments. An IDE (Integrated Developement Environment) for AEDL files may follow in one of the next
versions. The application software then uses this text file to initialize the experiment, where other secondary files might read for source fluxes, cross sections \etc . In this section, we show the definition of a very simple neutrino factory in AEDL, where we do not go into details. In the next chapter, we will discuss each of the individual steps in detail.

@@

Of course, the baseline and matter profile have to be specified, too, where
we use an arbitrary matter density profile here:

@@

{\tt \$profiletype = 3}\\

@@

@@

one element, because we only use one density layer: We initialize a baseline length of $3000 \, \mathrm{km}$ with a constant matter density of $3.5 \, \mathrm{g/cm^3}$. 

@@

The first element is the name of the flux, which we have defined above. 
The second element ``$\pm$'' determines whether 

@@

We now give a short introduction to the syntax of AEDL.
 The first eight characters have to be {\tt \%!GLoBES}
in order to avoid parsing megabytes of chunk
 and producing thousands of error messages. 
%
Comments can be used such as in C:
\begin{quote}
{\tt /* This starts a comment\\
 and here the comment ends */
}
\end{quote}
There are pre-defined variables which all start with {\tt \$}. Their range
is also checked. For example,  {\tt 
\$bins} can be only between $0$ and $500$.\footnote{The upper limit is 
only there for safety reasons, the memory is allocated dynamically.} If one uses a {\tt float} quantity where  an {\tt int} is expected, the {\tt float} will be converted to an {\tt int} in the same way as in C.  For example, we have scalar variables
\begin{quote}
{\tt
\$bins = 10\\
\$baseline = 1200.0
}
\end{quote}
and simple lists
\begin{quote}
{\tt
\$densitytab=\{1.0,2.2343,3.3432\} 
}
\end{quote}
%
Since there are often groups of data which we want to refer to later,
environments can be used. This is illustrated 
with the channel definition part:
\begin{quote}
{\tt channel(\#ch1)<\\
\tb  $\ldots$\\
>
}
\end{quote}
The first part is the type of environment, which is {\tt channel} here. 
There are the following types of environment in AEDL:
\begin{quote}
{\tt flux\\
cross\\
channel\\
energy\\
rule
}
\end{quote}
Besides the environment type, there is a user-defined name 
beginning with {\tt \#}
in the above example: {\tt \#ch1}. It can be used later to refer to the 
channel defined in {\tt <$\ldots$>}. Those names are so-called 
``automatic variables'' and have to start with {\tt \#}. Note that these names have to be unique and can only be refered to after their definition.
However, similar to C, one can give a header definition before:
\begin{quote}
{\tt    channel(\#ch2)<>}
\end{quote}
Now one can refer to the name {\tt \#ch2}, while the actual channel definition comes later. The internal representation of this automatic
variable is a number, which obtains its value from a counter for each type of environment. For example, for {\tt channel} the counter is {\tt numofchannels}. The counter keeps track of how many different names 
there are for one type of environment, which means that it counts the number of channels, rules, energy resolution functions \etc . Thus, the automatic
variables are numbered in the order of their definition, and the number
can later be used to refer to them in the C code (from $0$ to {\tt numof...}$-1$). Within each environment type, there are several 
variables beginning with {\tt @}, which can only be used within the 
appropriate type of environment. In many cases, 
they have a special syntax, such as {\tt @channel}.

\index{aedl}{NEXT@{\tt \#NEXT\#}}
If you want to have several experiments in one file, separate the different
 experiments by 
\begin{quote}
{\tt    \#NEXT\#}
\end{quote}
This command resets the counters for channels, rules, fluxes, and energy resolution environments. All variables have their scope limited 
by either {\tt \%!GLoBES, \#NEXT\#} or {\tt EOF}.  This allows 
a consistent treatment of various experiments in one file.

\index{aedl}{inlcude@{\tt include}}
As another feature of AEDL one can use include files with the {\tt include} command. Includes can be nested up to {\tt MAX\_INCLUSION\_DEPTH}, which is currently set to $10$. Error reporting works 
 for nested includes, too. The included file is not required to begin 
 with {\tt \%!GLoBES} to facilitate cut and paste:
\begin{quote}
{\tt include "./file\_1"}
\end{quote}
With this include mechanism, one can use constructions such as 
\begin{quote}
{\tt    include  "NuFact.gls"\\
        \#NEXT\#\\
        include   "JHFHK.gls"
}
\end{quote}
in order to initialize a combined analysis of the experiments defined in the files {\tt NuFact.gls} and {\tt JHFHK.gls}. Note that one has 
to use quotation marks for filenames in \AEDL.\index{norm}{File names}
Even if one uses the
automatic variable {\tt \#CC} in both experiments, 
but the cross section data are different (for example, because of different target nuclei), the correct 
cross section data will be applied to each of the experiments. 
Note that, alternatively, one can 
also load both files successively.

Furthermore, one can define constants such as
\begin{quote}
{\tt
Pi = 3.141
}
\end{quote}
These constants can not only be defined within one \AEDL\ file, but also
by the calling C program, which allows to use a simple but powerful variable
substitution mechanism as described in~\Sec~\ref{sec:aedlparams}.
\index{norm}{External \AEDL\ parameters}

In addition, some simple algebraic manipulations are possible, such as
\begin{quote}
{\tt
Pi+1\\
\verb+Pi^2+\\
sin(Pi/2)\\
}
\end{quote}
The following mathematical functions from {\tt <math.h>} are available: 
{\tt sin}, {\tt cos}, {\tt tan}, {\tt asin}, {\tt acos}, {\tt atan}, 
{\tt log}, {\tt log10}, {\tt exp}, {\tt sqrt}.
\index{aedl}{sin@{\tt sin}}
\index{aedl}{cos@{\tt cos}}
\index{aedl}{tan@{\tt tan}}
\index{aedl}{asin@{\tt asin}}
\index{aedl}{acos@{\tt acos}}
\index{aedl}{atan@{\tt atan}}
\index{aedl}{log@{\tt log}}
\index{aedl}{log10@{\tt log10}}
\index{aedl}{exp@{\tt exp}}
\index{aedl}{sqrt@{\tt sqrt}}

@@

\begin{table}[t!]
\begin{tabular}{|clp{7cm}|}
\hline
{\tt \$profiletype} & Additional variables & Description \\ 
\hline
$1$ & {\tt \$baseline} & Average density (constant) \\
$2$ & {\tt \$baseline}, {\tt \$densitysteps} & PREM profile with given number of equidistant steps \\
$3$ & {\tt \$lengthtab},  {\tt \$densitytab} & Arbitrary profile (table of layer thicknesses, table of densities) \\
\hline
\end{tabular}
\caption{\label{tab:profiletypes} Different matter density profiles which can be used with \GLOBES .}
\end{table}

@@

Furthermore, the matter density profile along the baseline
has to be specified. The simplest matter density profile is a constant matter density profile with the average matter density from the PREM\cite{Stacey} onion shell model of the Earth\index{norm}{PREM}\index{norm}{Earth matter density}:
\begin{quote}
{\tt \$profiletype=1 }
\end{quote}
%
For a better approximation of the realistic Earth matter density profile, one can use an arbitrary number of equidistant steps of the PREM profile:
\begin{quote}
{\tt \$profiletype=2 } \\
{\tt \$densitysteps=20 }
\end{quote}
Note that the value of {\tt \$densitysteps} is time-critical, since the
computation time of oscillation probabilities is directly 
proportional to the number of layers.
%
As a  third possibility, one can specify the matter density profile 
manually with a list of thicknesses and densities of the matter density layers. This example uses two density steps with two different densities:
\begin{quote}
{\tt \$profiletype=3 } \\
{\tt \$densitytab=\{2.8, 3.5\}}\\
{\tt \$lengthtab=\{1000.0, 2000.0\}}
\end{quote}
It is important that both lists habe the same length and that the  thicknesses given in  {\tt \$lengthtab} add up to the length of
the baseline, which does not have to be explicitely specified anymore. In addition, matter densities are always given in $g/cm^3$.
%
This approach can also be used for a constant matter density profile with
a specific matter density:
\begin{quote}
{\tt \$profiletype=3 } \\
{\tt \$densitytab=\{3.5\}}\\
{\tt \$lengthtab=\{3000.0\}}
\end{quote}
The possible options for matter density profiles are summarized in \Tab~\ref{tab:profiletypes}.

@@

A simple channel definition in \GLOBES\ consists of the flux,
the CP-sign of the initial state, the initial flavor, the final flavor,
the cross sections, and the energy resolution function. In order to refer to
flux, cross sections, and energy resolution functions, they have to be 
defined before with their {\tt \#name} in the respective environments. 
Thus, a simple definition of a channel\index{aedl}{{\tt channel}} is
\begin{quote}
{\tt channel(\#channel\_1)<\\
\tb @channel = \#flux : $+$: muon: muon: \#cross: \#energy\\
>}
\end{quote}
%
\index{norm}{Oscillation!switching off}
It is also possible to define a channel as no-oscillation by using the
prefix {\tt NOSC\_}\index{aedl}{{\tt channel}!{\tt NOSC\_}} 
in either the initial  flavour or 
the final flavour, like this
\begin{quote}
{\tt channel(\#channel\_1)<\\
\tb @channel = \#flux : $+$: NOSC\_muon: muon: \#cross: \#energy\\
>}
\end{quote}
%
In this case all diagonal probabilities $P_{\alpha\alpha}$ are unity, and all off-diagonal probablities  $P_{\alpha\beta}$ are zero. This is,
for instance, useful for
neutral current events, since these do not depend on any oscillation 
parameters\footnote{At least in the absence of sterile neutrinos}. The channels
marked as {\tt NOSC\_} are already computed by \GLB{SetRates}
and do not have to be recomputed in the subsequent fit (which calls the undocumented function \GLB{SetNewRates}).
Therefore this feature can be used to speed up the rate 
computation considerably, especially in cases where a large set
of channels exist which are only used for the computation of backgrounds.
Usually it is an excellent approximation to treat backgrounds as if they
were not affected by oscillations.

@@

In this section, we first discuss the principles of the energy smearing, where it is assumed that the reader is familiar with \Sec~\ref{sec:channel}. Then we introduce an automatic energy smearing algorithm, which is fairly simple to understand and applicable to most beam-based experiments. 
In most cases, the reader may want to continue to the next section after reading these two subsections. In the third subsection, we describe a more elaborate (and slower) smearing algorithm, which can be used together with rather fast neutrino oscillations compared to the bin size, such as for
(solar) reactor experiments to avoid aliasing effects. Eventually, we show how one can use a manual smearing matrix instead of using one of the implemented algorithms.

@@

Before we come to the calculation algorithms, it is useful to understand
the general evaluation algorithm. As it is illustrated in \figu{energies}, 
\GLOBES\ uses several levels with respect to the energy ranges:
\begin{description}
\item[Sampling point level]
 This level is used internally to evaluate the integrand in \eq~(\ref{eq:simple_int}) at all sampling points. The energy scale is the actual incident neutrino energy $E$. Note that the sampling range should exceed the energy bin range by about the energy resolution error, since the corresponding sampling points have to be evaluated by the algorithm. 
However, this exceeding is only
necessary to avoid aliasing effects if the integrand in \eq~(\ref{eq:simple_int}) is large at the limits of the sampling range.
For a manual definition of the sampling points (depends on algorithm), use
\index{aedl}{sampling points@{\tt \$sampling\_points}}
\index{aedl}{sampling min@{\tt \$sampling\_min}}
\index{aedl}{sampling max@{\tt \$sampling\_max}}
\begin{quote}
{\tt
\$sampling\_points = 20\\
\$sampling\_min =          4.0\\
\$sampling\_max =         50.0
}
\end{quote}
for equidistant sampling points. Arbitrarily spaced sampling points can 
be specified with 
{\tt \$sampling\_stepsize}\index{aedl}{sampling stepsize@{\tt \$sampling\_stepsize}}
\begin{quote}
{\tt
\$sampling\_stepsize=\{1.0,2.0,3.0,4.0,5.0,...\}
}
\end{quote}

@@

In order to ensure that fast oscillating probabilities do not lead to 
aliasing,\index{norm}{Aliasing}
it is possible to impose a low-pass filter already during the calculation
of the probabilities itself. This filter is implemented has a highly
experimental feature called ``filter''\index{norm}{Filter}. 
The calculation of oscillation
probabilities is, in principle, a computation of phase differences. Restricting the maximum admissible size of those phase differences effectively filters
the high frequency component of the oscillation probability. This idea is
implemented according to
\begin{eqnarray}
\label{eq:filter_a}
P_{\alpha\beta}(E)&=&\sum_{ij}
U_{\alpha j} U^*_{\beta j} U^*_{\alpha i} U_{\beta i} 
e^{-i\Phi_{ij}}\times 
e^{ -\Phi_{ij}^2/\sigma_f(E)^2 }\,,
\end{eqnarray}
where $\Phi_{ij}:=\Delta m_{ij}^2 L/2E$ is the usual phase difference and
the last term is a Gau\ss ian filter with width $\sigma_f(E)$. Choosing
$\sigma_f(E):=\sigma_f^0 \cdot E$ ensures that this filter behaves 
approximately such as an energy resolution function with constant width 
$\sigma_e=\sqrt{2}/\sigma_f^0$, \ie\
\begin{equation}
\label{eq:filter_b}
\int d\tilde E\quad P(\tilde E) \frac{1}{\sigma_e\,\sqrt{2\pi}}\,
e^{-\frac{(E-\tilde E)^2}{2\sigma^2_e}}\,.
\end{equation}
The relationship between \eqs~\ref{eq:filter_a} and~\ref{eq:filter_b}
is not obvious and connected to the properties of $P_{\alpha\beta}$: 
see \Refs~\cite{Kiers:1996zj,Giunti:2003ax}. This feature works \emph{only} 
for vacuum and constant densities and is controlled
by the filer state variable. In addition, $\sigma_e$ is set by the filter value variable:
\index{aedl}{filter state@{\tt \$filter\_state}}
\index{aedl}{filter value@{\tt \$filter\_value}}
\begin{quote}
{\tt
\$filter\_state = 1\\
\$filter\_value = 2.0\\
}
\end{quote}
would switch the filter feature on and set the width to $2.0\,\mathrm{GeV}$.
The setting of {\tt \$filter\_state} is ignored whenever a density profile
with more than one layer is used.
